Hierarchical Time Series Forecasting in Emergency Medical Services

### 2022-11-18 manuscript/main.tex
draft of cover letetr and title page

\usepackage{todonotes,mathtools}
\mathtoolsset{showonlyrefs}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Annals of Emergency Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical time series forecasting in Emergency Medical Services},
  pdfauthor={Bahman Rostami-Tabar; Rob J. Hyndman},
  pdfkeywords={forecasting, healthcare, emergency services, forecast
reconciliation, hierarchical time series, ambulance demand, attended
incidents},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Hierarchical time series forecasting in Emergency Medical
Services}
\author[1]{Bahman Rostami-Tabar%
\corref{cor1}%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[2]{Rob J. Hyndman%
%
}
 \ead{Rob.Hyndman@monash.edu}

@@

\begin{abstract}
Accurate forecasts of ambulance demand are crucial inputs when planning
and deploying staff and fleet. Such demand forecasts are required at
national, regional and sub-regional levels, and must take account of the
nature of incidents and their priorities. These forecasts are often
generated independently by different teams within the organization. As a
result, forecasts at different levels may be inconsistent, resulting in
conflicting decisions and a lack of coherent coordination in the
service. To address this issue, we exploit the hierarchical and grouped
structure of the demand time series, and apply forecast reconciliation
methods to generate both point and probabilistic forecasts that are
coherent and use all the available data at all levels of disaggregation.
The methods are applied to daily incident data from the Welsh Ambulance
Service NHS Trust, from October 2015 to July 2019, disaggregated by
nature of incident, priority, managing health board, and control area.
We use an ensemble of forecasting models, and show that the resulting
forecasts are better than any individual forecasting model. We validate
the forecasting approach using time-series cross-validation.
\end{abstract}

\begin{keyword}
    forecasting \sep healthcare \sep emergency services \sep forecast
reconciliation \sep hierarchical time series \sep ambulance demand \sep 
    attended incidents
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Inability to match the resources with the demand in Emergency Medical
Services (EMS) results in an overcrowding care system. This is a serious
problem causing challenging situations on patient flow with serious
consequences on patients, staff and the entire care system
\citep{ekstrom2015forecasting, ROSTAMITABAR20221197}. Demand forecasting
in EMS is a vital element that helps to depict various courses of action
to avoid the mismatch, which can result in massive savings in terms of
patient safety and lives. An accurate daily demand forecasting enables
planners and decision makers to manage resources to meet anticipated
patients, reconfigure units, redeploy staff and vehicles, where
necessary.

Demand forecasts at EMS are typically required at multiple levels of
cross-sectional granularities to inform various planning and
decision-making processes \citep{hulshof2012taxonomic}. There are some
planning process at the national level (more strategic or long-term)
such as workforce resource planning and budgeting; sub-national,
regional or healthcare level (tactical or medium-term) such as temporary
capacity expansions, resource sharing and staff-shift scheduling; and
hospital or station level (operational or short-term) such as planning
rosters for staff and ambulance deployment. Demand forecasts might also
be required at different level for a specific group of interest such as
nature of demand or priority. Moreover, the time series data in EMS has
an inherent hierarchical and grouped structure to support such
forecasting requirements. Demand for emergency medical services at the
country level can be disaggregated in a geographical hierarchy into
sub-national, regions, health boards, stations/hospitals or divided into
groups such as nature of incidents or demand priority. Therefore, using
forecasting methodologies that account for hierarchical and/or grouped
structures of time series in EMS seems to be a natural fit.

However, despite a large number of studies dedicated to forecasting for
EMS
\citep{mingliterature2022, gul2020exhaustive, ibrahim2016modeling, wargon2009systematic},
this area is neglected and the main focus has been on producing
forecasts at a single level, independently. Generating independent
forecasts not only ignore the inherent hierarchical and/or grouped
structure relationships of the time series demand but also results in a
lack of consistency and coordination. Consistent forecasts of the EMS
demand across all hierarchical and grouped levels are paramount for an
effective planning and decision making. The hierarchical forecasting
approaches can not only create consistent forecasts but also achieve
more accurate forecasts than the independent (base) forecasts
\citep{hyndman2011optimal}. Obtaining consistent forecasts at different
levels is important as it helps to avoid making conflicting decisions.
With hierarchical forecasting, plans at any level are based on coherent
forecasts and therefore can be aligned. Implementing and sustaining
improvements in EMS require alignments and coordination between
different stakeholders, without which teams operate in isolation leading
to conflicts, duplication work, rework, or work that runs counter to the
overall goal to improve the quality of delivery service. Hierarchical
forecasting framework can be used as a tool to improve coordination
between teams across the care services at the national, sub-national,
regional and local levels. To our knowledge, there is not only no
research in the EMS forecasting to account for the hierarchical and
grouped structure of the system but also in the entire field of
forecasting for healthcare management.

In this paper, we address this gap by investigating the application of
hierarchical forecasting approaches in the EMS using daily time series
of verified incidents from 2015 to 2020 in a major ambulance service in
Great Britain. The data has hierarchical and grouped structures with
hierarchies at the national, control (i.e.~sub-national), health board
(i.e.~regional) and groups by priority and nature of incidents. We
produce not only the point forecast but also the forecast distribution
across all levels, which is critical for an effective planning and
associated risk management. We compare the point and probabilistic
forecast accuracy of the independent forecasts, bottom-up and optimal
reconciliation approaches. We first generate independent/based forecasts
using Exponential Smoothing State Space (ETS), Generalized Linear Model
(GLM), Poisson regression, a simple empirical distribution and an
ensemble method followed by applying bottom-up and optimal
reconciliation approaches. Forecast performance is assessed by the Root
Mean Squared Scaled Error (RMSSE) for point forecasts and the Continuous
Ranked Probability Score (CRPS) for the probabilistic forecasts. This
paper complies with the principles of the reproducibility
\citep{stodden2013best, boylan2015reproducibility}. Therefore, the study
could equally be applied to any healthcare service (e.g.~emergency
department, primary or social care) subject to the time series having a
hierarchical and/or grouped structure, which is generally the case in
the healthcare sector.

The remainder of this article is structured as follows: In section 2, we
provide a brief review of the literature and discuss its limitation to
position our work; in section 3, we present the experiment design
describing the data set, forecasting methods and forecast evaluation
metrics. In Section 4, we discuss the hierarchical time series
forecasting approaches to generate both point and probabilistic
forecasts. In section 5, we present and discuss our results; in section
7, we summarize our findings and present ideas for future research.

\hypertarget{lit}{%
\section{Research background}\label{lit}}

Emergency medical services (EMS) are a critical component in the
delivery of urgent medical care to communities. An effective service
delivery requires accurate resource planning that are generally replying
on demand forecasts at operational, tactical and strategic levels.

There is a substantial number of studies on the application of time
series forecasting in the Emergency Medical Services. Various areas have
been the focus of the literature. Forecasting call volume arrivals is
one of the major research topics. \citet{ibrahim2016modeling} provide an
extensive review of the forecasting models in this context. Another
important area is related to forecasting ambulance demand. Although the
definition of demand might not be always clearly stated, however, this
is typically referring to a situation where a physical resource has been
deployed to respond to an incident. This might be also called
\emph{attended incidents}. Another demand related variable is verified
incidents. These are all incidents that require an action: either send a
physical vehicle, deal with via the Clinical Support Desk (e.g.~calls),
get an external (private) provider to respond to it, or send it through
to other channels such as police, firefighters or general practitioners.
Our study is aligned with this stream of the literature. Another similar
area that is largely studied in the literature, is Emergency Department
forecasting. We refer interested readers to \citet{mingliterature2022},
\citet{gul2020exhaustive} and \citet{wargon2009systematic} for extensive
reviews of the literature on Emergency Department forecasting. Although
crucial to EMS performance, \citet{aringhieri2017emergency} state that
demand forecasting has received limited research attention in the EMS
context. In this section, we provide a brief review of studies on
forecasting ambulance demand in EMS.

There are generally two main streams of research related to forecasting
ambulance demand in EMS: i) the first stream focuses on the application
of time series methods and regression approaches on forecasting
aggregate ambulance demand \citep{vile2012predicting, sasaki2010using};
ii) the second stream considers forecasting EMS demand in a more finer
temporal and geographical granularities by employing temporal-spatial
prediction methods \citep{zhou2016predicting, zhou2016predictinglit}.
The focus of our study is related to the first stream of research.

\citet{sasaki2010using} develop a multivariable regression model to
estimate future EMS demands. In addition to the historical demand, the
population census for different age groups and counts of the number of
companies employing more than five people are included in the
regression. The census variables describe groups who may be more likely
to need an ambulance. A stepwise ordinary least squares regression
analysis with SPSS is used for estimating the parameter and generating
forecast. The only performance measure reported in this study is
\(R^2\). The research design of this study is not rigorous and the study
is not reproducible. \citet{vile2012predicting} explore using a Singular
Spectrum Analysis (SSA) method to generate forecasts of the EMS demand
at the national level for 7-day, 14-day, 21-day and 28-day forecast
horizons using data provided by the Welsh Ambulance Service Trust
(WAST). The performance of this approach is compared with
Auto-Regressive Moving Average (ARIMA) and Holt-winter time series
methods using Root Mean Squared Error (RMSE). They concluded that point
forecasts generated by SSA are more accurate for longer-term, but ARIMA
and Holt-winter performance is superior for shorter-term horizons.
\citet{vile2016time} further develop a decision support system to
integrate forecasts generated by SSA. However, the study does not
compare and contrast the performance of forecasting methods based on
utility measures such as cost, resource utilization or response time.
The tool contains options that allow generating forecasts at various
levels of granularity, however, it ignore the hierarchical and grouped
relationships structure, preventing aligned decision making and
coordination.

\citet{9659837} investigate forecasting EMS demand in a high
spatio-temporal resolution of 1×1km spatial regions and 1-hr time
intervals using total incidents in Oslo, Norway from January 1st, 2015
up to and including February 11th, 2019. They used multi-layer
perceptron (MLP) and long short-term memory (LSTM) models to forecast
the EMS demand, and compare them to simple aggregation methods and
baselines. The point forecast accuracy is evaluated using Mean Absolute
Error (MAE) and Mean Squared Error (MSE) and the forecast distribution
is measures by Categorical Cross-Entropy. They shows that while Neural
Network models perform better in producing point forecast, a
distribution baseline method based on spatial distribution of the
incidents across all time steps provide more accurate forecast
distribution. \citet{zhou2016predictinglit} propose three methods based
on Gaussian mixture models, kernel density estimation, and kernel
warping to predict 4 weeks into future for a 1-km2 spatial region over
an hour. Two years of incidents attended from Toronto, Canada (years
2007 and 2008 with 391,296 events) and Melbourne, Australia (years 2011
and 2012 with 696,975 events) are used to build the model and examine
the performance on test data using mean negative log likelihood. They
show that forecasts generated by the proposed methods are significantly
more accurate than the current industry practice, a simple averaging
formula. \citet{grekousis2019will} investigate the combination of
spatial analysis methods with data mining techniques based on an
improved Hungarian algorithm and MLP neural network to identify the most
likely locations of future emergency events. The proposed approach is
tested using data of 2851 events attended by the EMS in Athens, Greece
over 24 weeks. They show that \(23.24\%\) of real emergency events lie
within 50 meter of the predicted ones and nearly \(70\%\) of the real
emergency events lie no further away than 150 meter, which is rather
accurate given the granularity of the problem at the city level.

We observed a number of limitations in the literature of EMS
forecasting, that encourage us to undertake this research. These
limitations are summarized as following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Current studies ignore the inherent hierarchical and/or grouped
  structure of the time series data and the relationship between series
  at different levels of hierarchy. This may result in incoherent
  forecasts leading to misaligned planning and decision making. While
  the hierarchical forecasting methodology has been developed and
  applied in various domains over the past 10 years
  \citep{panagiotelis2022probabilistic}, it has never been explored in
  this area.
\item
  Current research is mainly concerned with generating point forecast at
  a single level of hierarchy. There is a lack of studies presenting the
  entire forecast distribution of daily ambulance demand for the entire
  hierarchy to inform the whole decision-making process and to better
  represent the uncertainty of future demand, providing a risk
  management tool for planners.
\item
  Reproducibility is still a major challenge in EMS forecasting, as it
  is unlikely to reproduce the results without the help of the authors
  of those papers.
\item
  Another limitation is related to the generated forecasts that are not
  integer counts. Since actual ambulance counts cannot be negative or
  fractions, ambulance demand forecasts should be the same. While this
  might not be an issue when producing forecasts at a single level,
  producing non-negative count forecasts in a hierarchical/grouped
  structure is challenging and requires further investigation in the
  future.
\end{enumerate}

@@

We are interested in generating forecasts to inform the planning horizon
of \(ph= 42\) days, required by planners in the ambulance services
trust. The forecast horizon in this study is \(fh = 2 \times ph\) days
ahead (2*42 days planning horizon). This is because the planning is
generally frozen for \(ph\) days and considering a forecast horizon of
\(ph\) days might not be helpful for planning. While forecasts are
generated for \(2 \times ph\) days ahead, performance evaluation is only
assessed based on the last \(ph\) days and not the \(2 \times ph\) days.
The forecasts are produced for the holdout of 365 days using time series
cross-validation \citep{hyndman2021forecasting}.

In the following section, we discuss the dataset, describe the
forecasting methods used to generate base forecasts and present the
point and probabilistic accuracy measures.

@@

The dataset used in this study is from a major ambulance service trust
in Great Britain. It contains information relating to the daily number
of attended incidents from 1 October 2015 to 31 July 2019, disaggregated
by nature of incidents, priority, the health board managing the service
and the control area (or region). Figure~\ref{fig-hierarchy} depicts
both the hierarchical and grouped structure of the data.
Figure~\ref{fig-hierarchy-1} illustrates the nested hierarchical
structure based on control area and health board and
Figure~\ref{fig-hierarchy-2} shows the grouped structure by priority and
the nature of incident.

@@

Given the total number of time series, direct visual analysis is
infeasible. Therefore, we first compute features of all 1530 time series
\citep{m3pca} and display the strength of trend and weekly seasonality
strength in Figure~\ref{fig-feature}. Each point represents one time
series with the strength of trend in x-axis and the strength of
seasonality in y-axis. It is clear that there are some series showing
strong trends and/or seasonality, corresponding to series at the higher
levels of the hierarchy. The majority of series show low trend and
seasonality. These are time series belonging to the bottom series,
series related to the nature of incidents for a given control, health
board and priority level. Bottom series are dominated by noise with
little or no systematic patterns.

@@

\caption{\label{fig-dataviz2}Time series of attended incidents at
various levels.}

@@

Given the presence of various significant patterns in the past attended
incidents, we consider three different forecasting models to generate
the base forecasts. Once the base forecasts are produced, hierarchical
and grouped time series methods are used to reconcile them across the
all levels. We briefly discussed forecasting models in the following
sections, and the hierarchical forecasting methods are discussed in
Section~\ref{sec-htc}.

>>>>

\usepackage{todonotes,mathtools,bm,amsmath}
\mathtoolsset{showonlyrefs}
\usepackage{mathpazo}
\setlength{\parindent}{0cm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Annals of Emergency Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical Time Series Forecasting in Emergency Medical Services},
  pdfauthor={Bahman Rostami-Tabar; Rob J. Hyndman},
  pdfkeywords={healthcare, emergency services, forecast
reconciliation, ambulance demand, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\begin{document}

\begin{frontmatter}
\title{Hierarchical Time Series Forecasting in Emergency Medical
Services}
\author[1]{Bahman Rostami-Tabar%
\corref{cor1}%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[2]{Rob J. Hyndman%
%
}
 \ead{Rob.Hyndman@monash.edu}

@@

\begin{abstract}
Accurate forecasts of ambulance demand are crucial inputs when planning
and deploying staff and fleet. Such demand forecasts are required at
national, regional and sub-regional levels, and must take account of the
nature of incidents and their priorities. These forecasts are often
generated independently by different teams within the organization. As a
result, forecasts at different levels may be inconsistent, resulting in
conflicting decisions and a lack of coherent coordination in the
service. To address this issue, we exploit the hierarchical and grouped
structure of the demand time series, and apply forecast reconciliation
methods to generate both point and probabilistic forecasts that are
coherent and use all the available data at all levels of disaggregation.
The methods are applied to daily incident data from an ambulance service
in Great Britain, from October 2015 to July 2019, disaggregated by
nature of incident, priority, managing health board, and control area.
We use an ensemble of forecasting models, and show that the resulting
forecasts are better than any individual forecasting model. We validate
the forecasting approach using time-series cross-validation.
\end{abstract}

\begin{keyword}
    healthcare \sep emergency services \sep forecast
reconciliation \sep ambulance demand \sep 
    regression
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

A failure to match available resources to demand in Emergency Medical
Services (EMS) results in patient flow problems, with serious
consequences for patients, staff and the entire care system
\citep{ekstrom2015forecasting, ROSTAMITABAR20221197}. Demand forecasting
in EMS helps service planners to avoid the mismatch, potentially
providing massive savings in costs and lives, and leading to better
patient outcomes. Accurate daily demand forecasting enables planners and
decision makers to manage resources to meet anticipated patients,
reconfigure units, and redeploy staff and vehicles as necessary.

Demand forecasts at EMS are typically required at multiple levels of an
organization to inform various planning and decision-making processes
\citep{hulshof2012taxonomic}. There are some planning process at the
national level (strategic and long-term) such as workforce resource
planning and budgeting; sub-national, regional or healthcare level
(tactical and medium-term) such as temporary capacity expansions,
resource sharing and staff-shift scheduling; and hospital or station
level (operational and short-term) such as planning rosters for staff
and ambulance deployment. Demand forecasts might also be required at
different levels for a specific area of interest such as the nature of
demand or the priority level. Moreover, the time series data in EMS has
an inherent hierarchical and grouped structure to support such
forecasting requirements. Demand for emergency medical services at the
country level can be disaggregated in a geographical hierarchy into
sub-national, regions, health boards, and stations/hospitals, or divided
into groups such as the nature of incidents or demand priority.
Therefore, using forecasting methodologies that account for hierarchical
and/or grouped structures of time series in EMS is a natural fit.

However, despite a large number of studies dedicated to forecasting for
EMS
\citep{mingliterature2022, gul2020exhaustive, ibrahim2016modeling, wargon2009systematic},
the hierarchical data structure has been largely ignored, and the main
focus has been on producing independent forecasts at a single level.
Generating independent forecasts can result in a lack of consistency and
coordination, and therefore leads to less effective planning and
decision making. With hierarchical forecasting, plans at any level are
based on coherent forecasts and therefore can be aligned. Implementing
and sustaining improvements in EMS require alignments and coordination
between different stakeholders, without which teams operate in isolation
leading to conflicts, duplication work, rework, or work that runs
counter to the overall goal to improve the quality of delivery service.
Hierarchical forecasting framework can be used as a tool to improve
coordination between teams across the care services at the national,
sub-national, regional and local levels. The hierarchical forecasting
approaches not only create consistent forecasts, but are usually also
more accurate than the independent (base) forecasts
\citep{hyndman2011optimal}. To our knowledge, there has been no previous
research involving hierarchical and grouped forecasting in the entire
field of forecasting for healthcare management.

In this paper, we address this gap by investigating the application of
hierarchical forecasting approaches in the EMS using daily time series
of verified incidents from 2015 to 2020 in a major ambulance service in
Great Britain. The data has hierarchical and grouped structures, with
hierarchies at the national, control (i.e.~sub-national), health board
(i.e.~regional) levels, as well as groups by priority and nature of
incidents. We produce consistent point forecasts and forecast
distributions for all levels, which is critical for an effective
planning and associated risk management. We compare the point and
probabilistic forecast accuracy of the independent forecasts, bottom-up
and optimal reconciliation approaches. We first generate
independent/base forecasts using Exponential Smoothing State Space
(ETS), Generalized Linear Model (GLM), Poisson regression, a simple
empirical distribution and an ensemble method, followed by applying
bottom-up and optimal reconciliation approaches. Forecast performance is
assessed by the Mean Squared Scaled Error (MSSE) for point forecasts and
Continuous Ranked Probability Scores (CRPS) for the probabilistic
forecasts. This paper complies with reproducibility principles
\citep{stodden2013best, boylan2015reproducibility}, and can be applied
to any healthcare service (e.g., emergency department, primary or social
care) subject to the time series having a hierarchical and/or grouped
structure.

The remainder of this article is structured as follows: In
Section~\ref{sec-lit}, we provide a brief review of the literature and
discuss its limitation to position our work; in
Section~\ref{sec-experiment}, we present the experiment design
describing the data set, forecasting methods and forecast evaluation
metrics. In Section~\ref{sec-htc}, we discuss the hierarchical time
series forecasting approaches to generate both point and probabilistic
forecasts. In Section~\ref{sec-results}, we present and discuss our
results; in Section~\ref{sec-conclusion}, we summarize our findings and
present ideas for future research.

\hypertarget{sec-lit}{%
\section{Research background}\label{sec-lit}}

Emergency medical services (EMS) are a critical component in the
delivery of urgent medical care to communities. An effective service
delivery requires accurate resource planning that generally relies on
demand forecasts at operational, tactical and strategic levels.

There is a substantial number of studies on the application of time
series forecasting in the Emergency Medical Services. For example,
\citet{ibrahim2016modeling} provide an extensive review of the models
used in forecasting call volume arrivals. Another important area is
related to forecasting ambulance demand. Although the definition of
demand might not be always clearly stated, this is typically referring
to a situation where a physical resource has been deployed to respond to
an incident. This might be also called \emph{attended incidents}.
Another demand related variable is verified incidents; these are all
incidents that require an action: either by sending a physical vehicle,
responding via the Clinical Support Desk, requesting an external
provider to respond to it, or forwarding it to other channels such as
police, firefighters or general practitioners. Our study is aligned with
this stream of the literature. Another similar area that has received
considerable attention is Emergency Department forecasting; we refer
interested readers to \citet{mingliterature2022},
\citet{gul2020exhaustive} and \citet{wargon2009systematic} for extensive
reviews of the relevant literature. Although crucial to EMS performance,
\citet{aringhieri2017emergency} state that demand forecasting has
received limited research attention in the EMS context. In this section,
we provide a brief review of studies on forecasting ambulance demand in
EMS.

There are generally two main streams of research related to forecasting
ambulance demand in EMS: (i) the first stream focuses on the application
of time series methods and regression approaches to forecasting
aggregate ambulance demand \citep{vile2012predicting, sasaki2010using};
and (ii) the second stream considers forecasting EMS demand in finer
temporal and geographical granularities by employing temporal-spatial
prediction methods \citep{zhou2016predicting, zhou2016predictinglit}.
The focus of our study is related to the first stream of research.

\citet{sasaki2010using} develop a multivariable regression model to
estimate future EMS demands. In addition to the historical demand, the
population census for different age groups and counts of the number of
companies employing more than five people are included in the
regression. The census variables describe groups who are more likely to
need an ambulance. A stepwise ordinary least squares regression analysis
is used for estimating the parameter and generating forecast. The only
performance measure reported in this study is \(R^2\), which is not an
effective measure of forecast accuracy \citep[p457]{armstrong01}. The
research design of this study is not rigorous and the study is not
reproducible. \citet{vile2012predicting} explore using a Singular
Spectrum Analysis (SSA) method to generate forecasts of the EMS demand
at the national level for 7-day, 14-day, 21-day and 28-day forecast
horizons using data provided by an ambulance service in Great Britain.
The performance of this approach is compared to Auto-Regressive
Integrated Moving Average (ARIMA) and Holt-Winters time series methods
using Root Mean Squared Error (RMSE). They concluded that point
forecasts generated by SSA are more accurate for longer-term, but that
ARIMA and Holt-Winters performance is superior for shorter-term
horizons. \citet{vile2016time} further develop a decision support system
to integrate forecasts generated by SSA. However, the study does not
compare and contrast the performance of forecasting methods based on
utility measures such as cost, resource utilization or response time.
The tool contains options that allow generating forecasts at various
levels of granularity, however, it ignores the hierarchical and grouped
relationships structure, preventing aligned decision making and
coordination.

\citet{9659837} investigate forecasting EMS demand in a high
spatio-temporal resolution of 1km\(^2\) spatial regions and 1-hr time
intervals using total incidents in Oslo, Norway, from 1 January 2015 to
11 February 2019. They used multi-layer perceptron (MLP) and long
short-term memory (LSTM) models to forecast the EMS demand, and compare
the results to simple aggregation methods and baselines. The point
forecast accuracy is evaluated using Mean Absolute Error (MAE) and Mean
Squared Error (MSE), and the forecast distribution is measured by
Categorical Cross-Entropy. They found that Neural Network models
performed better in producing point forecasts, while a distribution
baseline method based on spatial distribution of the incidents across
all time steps provided more accurate forecast distributions.
\citet{zhou2016predictinglit} proposed three methods based on Gaussian
mixture models, kernel density estimation, and kernel warping to predict
hourly data 4 weeks ahead for a 1km\(^2\) spatial region. Two years of
incidents from Toronto, Canada (years 2007 and 2008 with 391,296 events)
and Melbourne, Australia (years 2011 and 2012 with 696,975 events) were
used to build the model and examine the performance on test data using
mean negative log likelihood. They show that forecasts generated by the
proposed methods were significantly more accurate than the current
industry practice (a simple averaging formula).
\citet{grekousis2019will} investigated the combination of spatial
analysis methods with data mining techniques based on an improved
Hungarian algorithm and a MLP neural network to identify the most likely
locations of future emergency events. The proposed approach was tested
using data of 2851 events attended by the EMS in Athens, Greece, over 24
weeks. They showed that 23\% of real emergency events lie within 50
meters of the predicted ones and nearly 70\% of the real emergency
events lie no further than 150 meters away, which is rather accurate
given the granularity of the problem at the city level.

We note a number of limitations in the literature of EMS forecasting,
that encourage us to undertake this research. These limitations are
summarized as following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Current studies ignore the inherent hierarchical and/or grouped
  structure of the time series data, and the relationship between series
  at different levels of hierarchy. This may result in incoherent
  forecasts leading to misaligned planning and decision making. While
  the hierarchical forecasting methodology has been developed and
  applied in various domains over the past 10 years
  \citep{panagiotelis2022probabilistic}, it has never been explored in
  this area.
\item
  Current research is mainly concerned with generating point forecasts
  at a single level of hierarchy. There is a lack of studies considering
  the entire forecast distribution of daily ambulance demand for the
  whole hierarchy to inform the decision-making process and to better
  represent the uncertainty of future demand, providing a risk
  management tool for planners.
\item
  Reproducibility is still a major challenge in EMS forecasting, as it
  is unlikely that any reader can reproduce prior studies without the
  help of the authors of those papers.
\item
  Another limitation is related to the generated forecasts not being on
  the sample space of non-negative counts. Since actual ambulance counts
  cannot be negative or non-integer, ambulance demand forecast
  distributions should reflect the data. Of course, point forecasts
  represent means, so they should be non-negative, but may be
  non-integer. While this might not be an issue when producing forecasts
  at a single level, producing non-negative count forecasts in a
  hierarchical/grouped structure is challenging and requires further
  investigation in the future.
\end{enumerate}

@@

We are interested in generating forecasts to inform the planning horizon
of 42 days, required by planners in the ambulance service. The forecast
horizon in this study is \(2 \times 42 = 84\) days ahead, because the
planning is generally frozen for 42 days and so forecasts of the next 42
days is not particularly helpful for planning. While forecasts are
generated for 84 days ahead, performance evaluation is only assessed
based on the last 42 days and not the whole forecast period. The
forecasts are produced for various training and test sets using time
series cross-validation \citep{hyndman2021forecasting}.

In the following section, we discuss the dataset, describe the
forecasting methods used to generate base forecasts, and present the
point and probabilistic accuracy measures.

@@

The dataset used in this study is from a major ambulance service in
Great Britain. It contains information relating to the daily number of
attended incidents from 1 October 2015 to 31 July 2019, disaggregated by
nature of incidents, priority, the health board managing the service and
the control area (or region). Figure~\ref{fig-hierarchy} depicts both
the hierarchical and grouped structure of the data.
Figure~\ref{fig-hierarchy-1} illustrates the nested hierarchical
structure based on control area and health board and
Figure~\ref{fig-hierarchy-2} shows the grouped structure by priority and
the nature of incident.

@@

Given the total number of time series, direct visual analysis is
infeasible. Therefore, we first compute features of all 1530 time series
\citep{m3pca} and display the strength of trend and weekly seasonality
strength in Figure~\ref{fig-feature}. Each point represents one time
series with the strength of trend in x-axis and the strength of
seasonality in y-axis. Both measures are on a scale of {[}0,1{]}. It is
clear that there are some series showing strong trends and/or
seasonality, corresponding to series at the higher levels of the
hierarchy. The majority of series show low trend and seasonality. These
are time series belonging to the bottom series, series related to the
nature of incidents for a given control, health board and priority
level. Bottom series are dominated by noise with little or no systematic
patterns.

@@

\caption{\label{fig-dataviz2}Time series of attended incidents at
various levels. The panels show data from the whole country, by control
area, by health board, by priority level, and by nature of incident.
Only four of the 35 nature of incident categories are shown to avoid too
much overplotting.}

@@

Given the presence of various significant patterns in the past attended
incidents, we consider three different forecasting models to generate
the base forecasts. Once the base forecasts are produced, hierarchical
and grouped time series methods are used to reconcile them across all
levels. We briefly discuss forecasting models in the following sections,
and the hierarchical forecasting methods are discussed in
Section~\ref{sec-htc}.

\textbf{Naive:} We start with a simple forecasting approach, assuming
that the future days will be similar to past days. We use the empirical
distribution of the past daily attended incidents to create the forecast
distribution of future attended incidents.
